---
title: "STA106 Project 3"
author: "Amir Yaacoobi"
date: "2023-03-19"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(42)
library(iNZightTools)
library(DescTools)
heart <- read.csv("~/Desktop/heart_disease_health_indicators_BRFSS2015.csv")
heart$Smoker=as.factor(heart$Smoker)
heart$HighBP=as.factor(heart$HighBP)
heart$HighChol=as.factor(heart$HighChol)

Build_contigencytable=function(data,group,variable,bins=10,proportion=FALSE){
  table1=NULL
  # create break points for the following histograms
  # from minimum to maximum with equal distance
  ax=seq(min(data[,variable]),max(data[,variable]),length.out=bins+1)
  # Save histogram data
  list_group=unique(data[,group])
  for(i in list_group){
    hg1=hist(data[data[,group]==i,variable], breaks = ax,plot = FALSE)
    table1=rbind(table1,hg1$counts)
  }
  rownames(table1)=list_group
  colnames(table1)=1:ncol(table1)
  # calculate row sum and combine it  with the current table
  table1=cbind(table1, 'Total'=apply(table1,1,sum))
  # calculate column sum and combine it  with the current table
  table1=rbind(table1, 'Total'=apply(table1,2,sum))

  if(proportion){
    # convert to proportions
    n_col=ncol(table1)
    for(i in 1:nrow(table1)){
      table1[i,]=table1[i,]/table1[i,n_col]
    }
  }
  table1
}
```
```{r}
entropyApproach = function(sample1){
  B=nrow(sample1)
  entropies=numeric(B)
  for(j in 1:B){
    entropies[j]= Entropy(sample1[j,],base=exp(1))
  }
    labels = row.names((sample1))
    entropiesTable = cbind(labels, entropies)
    return(entropiesTable)
}


hcApproach = function(propTable){
  varClusters = hclust(dist(propTable), method='average')
  plot(varClusters, main="Dendogram of Proportions")
}

```

```{r}

entropyHistSim = function(table1){
for (i in 1:nrow(table1)){
  sample1=rmultinom(1000,100,table1[i,])
  B=ncol(sample1)
  entropies=numeric(B)
  for(j in 1:B){
    entropies[j]=Entropy(sample1[,j],base=exp(1))
  }
  hist(entropies,main='Histograms of Entropy',xlab=paste('Entropy for', row.names(table1)[i]))
}
}
##^repeat that for all 2-way interaction terms

```
```{r}
hcApproachSim = function(numsTable, propTable){
  table1 = numsTable
  table2 = propTable
  simTable = NULL
    for (i in 1:(nrow(table1)-1)){
    sample1=rmultinom(n=1, size=table1[i, ncol(table1)],prob=table2[i,1:ncol(table2)-1])
    simTable = rbind(simTable, t(sample1))
    }
  rownames(simTable) = rownames((table2[1:nrow(table2)-1,]))
  simTable=cbind(simTable, 'Total'=apply(simTable,1,sum))
  simTable=rbind(simTable, 'Total'=apply(simTable,2,sum))
  n_col=ncol(simTable)
    for(i in 1:nrow(simTable)){
      simTable[i,]=simTable[i,]/simTable[i,n_col]
    }
  varClusters = hclust(dist(simTable), method='average')
  plot(varClusters, main="Dendogram of Proportions")
}
```



### ABSTRACT
We have been given the BRFSS 2015 data set and told to explore the data through the use of entropy and HC-clustering with respect to three binary variables of our choice, subdivided in to five datasets with respect to general health. We have found that in terms of predictabilty, the interaction of sex with blood pressure and cholesterol status seemed to be most impactful, with blood pressure playing a secondary role. Additionally, through repeated simulated samples using the multinomial distribution, we found that the use of entropy and HC-trees were very reliable, consistently generating sample data that matched the actual entropy and HC-trees generated from our dataset.

### INTRODUCTION
We have been provided with the BRFSS 2015 data set and introduced to the methods of hierarchical-clustering and Entropy as alternatives to the method of ANOVA in the task of determining whether or not multiple samples have the same mean or distribution. 

Now, we are going to put apply these methods to our BRFSS dataset by dividing our dataset into five samples with regard to the GenHlth category, and investigating the 3-way interaction effects of three variables: sex, high cholesterol status, and high blood pressure status. Using these methods of hierarchical-clustering and entropy, we will determine what association the interactions of sex, high cholesterol, and high blood pressure, has on BMI distribution, and how it compares to the overall BMI distribution.

However, we will also test the reliability of entropy and HC-trees through the use of simulated data based on the multinomial distribution for each sub-dataset with respect to 3-way interactions.

### METHODOLOGY

Each of our sub-datasets with regard to GenHlth will be further subdivided with respect to high blood pressure, high cholesterol, and sex. These datasets will be represented through contingency tables and contingecy tables of proportions. Then, through the use of hierarchical clustering we will discover those with similar distributions, and with entropy we will see which combination of variables will have the greatest predictor of BMI.

To test reliability, a contingency table of proportions will be constructed for each 3-way interaction sub-dataset. Then, we will simulate data for each contingency table from a multinomial distribution in which each row-vector of proportions will serve as the probability for the multinomial distribution. 

For the entropy approach, each row will construct its own multinomial distribution with n=1000. Then we use the entropy approach for each row to create a histogram of the Shannon entropies and compare to the actual data.

For the hierarchical clustering approach, we will create 5 such contingency tables as detailed above, and construct HC-trees on each table to investigate how reliable HC is.

```{r}
combined = combineCatVars(heart, vars=c("Smoker", "HighBP", "HighChol"), sep=".")
combined = combineCatVars(combined, vars=c("HighBP", "HighChol"), sep=".")
combined = combineCatVars(combined, vars=c("Smoker", "HighChol"), sep=".")
combined = combineCatVars(combined, vars=c("Smoker", "HighBP"), sep=".")
```

### KEY RESULTS

##### SUB-DATASET 1

```{r}
sub=combined[combined$GenHlth==1,]
subTotTable = Build_contigencytable(sub, "Smoker.HighBP.HighChol", "BMI", 10, proportion=F)
subPropTable = Build_contigencytable(sub, "Smoker.HighBP.HighChol", "BMI", 10, proportion=T)
entropyApproach(subPropTable)
```

```{r}
hcApproach(subPropTable)
```

```{r fig.height=7, fig.width=14}
par(mfrow=c(3, 3))
entropyHistSim(subPropTable)
par(mfrow=c(3, 2))
for (i in 1:5){
  hcApproachSim(subTotTable, subPropTable)
}
```

##### SUB-DATASET 2

```{r}
sub=combined[combined$GenHlth==2,]
subTotTable = Build_contigencytable(sub, "Smoker.HighBP.HighChol", "BMI", 10, proportion=F)
subPropTable = Build_contigencytable(sub, "Smoker.HighBP.HighChol", "BMI", 10, proportion=T)
entropyApproach(subPropTable)
```

```{r}
hcApproach(subPropTable)
```

##### RELIABILITY
```{r fig.height=7, fig.width=14}
par(mfrow=c(3, 3))
entropyHistSim(subPropTable)
par(mfrow=c(3, 2))
for (i in 1:5){
  hcApproachSim(subTotTable, subPropTable)
}
```

##### SUB-DATASET 3
```{r}
sub=combined[combined$GenHlth==3,]
subTotTable = Build_contigencytable(sub, "Smoker.HighBP.HighChol", "BMI", 10, proportion=F)
subPropTable = Build_contigencytable(sub, "Smoker.HighBP.HighChol", "BMI", 10, proportion=T)

entropyApproach(subPropTable)
```

```{r}
hcApproach(subPropTable)
```

##### RELIABILITY

```{r fig.height=7, fig.width=14}
par(mfrow=c(3, 3))
entropyHistSim(subPropTable)
par(mfrow=c(3, 2))
for (i in 1:5){
  hcApproachSim(subTotTable, subPropTable)
}
```

##### SUB-DATASET 4
```{r}
sub=combined[combined$GenHlth==4,]
subTotTable = Build_contigencytable(sub, "Smoker.HighBP.HighChol", "BMI", 10, proportion=F)
subPropTable = Build_contigencytable(sub, "Smoker.HighBP.HighChol", "BMI", 10, proportion=T)

entropyApproach(subPropTable)
```

```{r}
hcApproach(subPropTable)
```

##### RELIABILITY
```{r fig.height=7, fig.width=14}
par(mfrow=c(3, 3))
entropyHistSim(subPropTable)
par(mfrow=c(3, 2))
for (i in 1:5){
  hcApproachSim(subTotTable, subPropTable)
}
```

##### SUB-DATASET 5
```{r}
sub=combined[combined$GenHlth==5,]
subTotTable = Build_contigencytable(sub, "Smoker.HighBP.HighChol", "BMI", 10, proportion=F)
subPropTable = Build_contigencytable(sub, "Smoker.HighBP.HighChol", "BMI", 10, proportion=T)
entropyApproach(subPropTable)
```


```{r}
hcApproach(subPropTable)
```

##### RELIABILITY

```{r fig.height=7, fig.width=14}
par(mfrow=c(3, 3))
entropyHistSim(subPropTable)
par(mfrow=c(3, 2))
for (i in 1:5){
  hcApproachSim(subTotTable, subPropTable)
}
```

